flowCore: A Bioconductor software package for high throughput flow cytometry analysis

### 2008-08-15 paper/manuscript.tex
git-svn-id: https://hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@33255 bc3139a8-67e5-0310-9ffc-ced21a209358

\section*{Introduction}
\subsection*{Computational tools for the analysis of FC-HCS data}
During the last several years, automation technologies have been
developed that enable the use of FCM high content screening
(FC-HCS) to generate large, complex datasets in both basic and
clinical research applications \citep{brinkman2007hcf}. A serious
bottleneck in the interpretation of existing studies and the
application of FC-HCS to even larger, more complex problems is that
data management and data analysis methods have not advanced
sufficiently far from the methods developed for applications of flow
cytometry (FCM) to small-scale, tube-based studies.

Some of the consequences of this lag in development are difficulties
in maintaining the integrity and documentation of extremely large
datasets, assessing measurement quality, developing validated assays,
controlling the accuracy of gating techniques, automating complex
gating strategies, and aggregating statistical results across large
study sets for further analysis.  Among the most serious consequences
of the current situation is, however, that it is very difficult for
new analysis approaches to find their way into standard practice. We
believe that this barrier to the development and dissemination of new
analysis methods is one of the fundamental restraints on the future
expansion of FC-HCS in both clinical and research applications.

In this paper, we describe a set of flexible and well structured
computational tools to efficiently analyze FC-HCS. This set of tools
is based on the open source statistical software R \citep{Rmain}, in
conjunction with the Bioconductor Project \citep{BIOC}. Our intent is
to provide a shared research platform that enables bioinformaticians,
computer scientists, and statisticians to work collaboratively with
biologists and clinicians to develop novel methods for FCM data
analysis, a process deemed crucial by many for the further development
of the technology \citep{lizard2007fca}. A critical component of this
approach is the development and implementation of standards that will
facilitate the adoption of these methods by both the larger research
community and commercial software vendors.  Consequently, we are
optimistic that this software platform will play a critical role in
addressing the bottlenecks described above that limit further advances
in the application of FC-HCS.

The computational tools we have developed are distributed in the R
software language as the Bioconductor package
\Rpackage{flowCore}. \Rpackage{flowCore} is a freely available, highly
functional and extensible FCM data analysis platform that enables
researchers to efficiently handle FC-HCS data and encourages open
development of tools for their coherent analysis. \Rpackage{flowCore}
implements computationally efficient data structures and a range of
specialized methods addressing all components of a typical FCM
analysis work flow, including compensation, transformation, and
gating. \Rpackage{flowCore} runs on Windows, Mac OS X and Linux/Unix
operating systems. Conceptually, the underlying data structures that
build the core of our software can be implemented in any programming
language. We chose the R language for its advantages in statistical
programming and visualization. In this paper, we describe the
functionality and methods underlying \Rpackage{flowCore}, and provide
examples of their use.

>>>>

\section*{Introduction}
\subsection*{Computational tools for the analysis of FC-HCS data}
Automation technologies developed during the last several years have
enabled the use of FCM high content screening (FC-HCS) to generate
large, complex datasets in both basic and clinical research
applications \citep{brinkman2007hcf}. A serious bottleneck in the
interpretation of existing studies and the application of FC-HCS to
even larger, more complex problems is that data management and data
analysis methods have not advanced sufficiently far from the methods
developed for applications of flow cytometry (FCM) to small-scale,
tube-based studies.

Some of the consequences of this lag are difficulties in maintaining
the integrity and documentation of extremely large datasets, assessing
measurement quality, developing validated assays, controlling the
accuracy of gating techniques, automating complex gating strategies,
and aggregating statistical results across large study sets for
further analysis. The most serious consequence, however, is that new
analysis approaches face difficulty in finding their way into standard
practice. We believe that this barrier to the development and
dissemination of new analysis methods is one of the fundamental
restraints on the future expansion of FC-HCS in both clinical and
research applications.

In this paper, we describe a set of flexible and well structured
computational tools to efficiently analyze FC-HCS data. This set of
tools is based on the open source statistical software R
\citep{Rmain}, in conjunction with the Bioconductor Project
\citep{BIOC}. Our intent is to provide a shared research platform that
enables bioinformaticians, computer scientists, and statisticians to
work collaboratively with biologists and clinicians to develop novel
methods for FCM data analysis, a process deemed crucial by many for
the further development of the technology \citep{lizard2007fca}. A
critical component of this approach is the development and
implementation of standards that will facilitate the adoption of these
methods by both the larger research community and commercial software
vendors.  Consequently, we are optimistic that this software platform
will play a critical role in addressing the bottlenecks described
above that limit advances in the application of FC-HCS.

The computational tools we have developed are distributed in the R
software language as the Bioconductor package
\Rpackage{flowCore}. \Rpackage{flowCore} is a freely available, highly
functional, and extensible FCM data analysis platform that enables
researchers to efficiently handle FC-HCS data and encourages open
development of tools for their coherent analysis. \Rpackage{flowCore}
implements computationally efficient data structures and a range of
specialized methods addressing all components of a typical FCM
analysis work flow, including compensation, transformation, and
gating. \Rpackage{flowCore} runs on Windows, Mac OS X, and Linux/Unix
operating systems. Conceptually, the underlying data structures that
build the core of our software can be implemented in any programming
language. We chose the R language for its advantages in statistical
programming and visualization. In this paper, we describe the
functionality and methods underlying \Rpackage{flowCore}, and provide
examples of their use.
